Abstractions for Multi-Sorted Substitutions

### 2023-09-20 paper/paper.tex
Update.

  % TODO: avoiding interaction lemmas

@@

  The rest of this paper is structured as followed:
  Section~\ref{sec:syntax} introduces the multi-sorted syntax and compares it to the
  more common unsorted syntax.
  Section~\ref{sec:maps} introduces multi-sorted substitutions and renamings, and
  an abstraction to unify them.
  Section~\ref{sec:composition} introduces composition of multi-sorted
  substitions and renamings, and an abstraction to unify all four
  compositions.
  Section~\ref{sec:types} shows how to introduce the notion of types
  for multi-sorted syntax and how to define typing contexts.
  Section~\ref{sec:typing} shows how to define a multi-sorted typing relation,
  and an abstraction to unify type preservation lemmas for renamings and substitutions.
  Section~\ref{sec:subjectreduction} introduces a small-step
  operational semantics for System F and proves subject reduction.
  Section~\ref{sec:reflection} describes the class of object languages covered by
  our reflection algorithm.
  Section~\ref{sec:casestudies} discusses our case studies.
  Section~\ref{sec:related} discusses related work.
  Section~\ref{sec:future} discusses future work.
  Section~\ref{sec:conclusion} concludes.

@@

  % TODO note that its derived

@@

  Frameworks for programming language metatheory can be broadly
  categorized by representation of variable bindings and
  representation of typing and scoping.

  Variable bindings have been represented as de Bruijn indices, de
  Bruijn levels, names, locally nameless, locally named, shifted names,
  nominal techniques, (parameterized) higher order abstract syntax.

  Typing and scoping can be both extrinsic, both intrinsic, or
  extrinsically typed and intrinsically scoped.

  As the amount of related work is rather large, we focus on work that
  is more closely related to ours and refer to other papers for the
  broader picture. In particular, we discuss abstractions for renamings
  and substitutions, comparable extrinsically typed approaches, but also
  intrinsically typed approaches that could be used as a basis for
  multi-sorted syntax, which as mentioned before is intrinsically typed
  in disguise.

@@

  The kit abstraction for unifying renamings and substitutions
  appeared first in an unpublished manuscript by
  McBride\cite{unpublished:mcbride2005kits}, and later in
  Benton et al.\cite{DBLP:journals/jar/BentonHKM12}. 
  Wood and Atkey\cite{DBLP:journals/corr/abs-2005-02247} propose an
  extension to kits that supports substructural types via resource
  vectors.
  In all three cases kits are formulated for intrinsic typing
  and scenarios with polymorphism are not considered.

@@

  \emph{Autosubst 2}\cite{DBLP:conf/cpp/StarkSK19} is a standalone
  code generator, which translates second-order HOAS specifications into
  mutual inductive term sorts. Compared to Autosubst 1, it features
  mutually recursive object languages, intrinsically scoping, and
  vectorized substitutions.
  They also define and prove lemmas for all four compositions.
  The generated syntax consists of multiple term types, indexed by
  different amounts of free variables as shown in Section~\ref{sec:syntax}.

  Both works do not provide machinery for type preservation and
  do not unify renaming and substitution (although the latter is listed as
  future work). We believe it should be possible to translate their
  decision procedure from vectorized substitutions to the setting of
  our multi-sorted substitutions.

  Abstract binding trees

@@

  Allais et al.\cite{DBLP:conf/cpp/Allais0MM17} propose a powerful
  abstraction for denotational semantics and semantic fusion lemmas.
  In later work\cite{DBLP:journals/pacmpl/AllaisA0MM18}, they use
  generic programming to instantiate this abstraction for a class of
  object languages comparable to ours.
  They demonstrate how both renamings and substitutions can be
  described as semantics and how the four composition lemmas follow from
  their generic fusion lemma.
  They do cover an abstraction to unify renamings and substitutions.
  They do not show how to work with multiple variable sorts, how to
  abstract over the compositions, and how to abstract over typing in
  an extrinsically typed scenario.
  However, we believe that their framework is general enough such that
  our concepts can also be transferred to their framework.

  This work is orthogonal but compatible to our approach.
  We can instantiate the system using our syntax type system,
  but then the semantics abstraction would work with terms
  which are untyped with respect to the actual type system.
  We think it is worthwile to explore if their semantics abstraction
  can also be generalized to multi-sorted typing relations.

  \section{Future Work}
  \label{sec:future}
  \begin{itemize}
  \item
    Enabling more object languages by allowing subterms with less
    restrictions on the form of \ACode{S}, allowing recursive
    occurences in container types like lists and finite functions, and
    adding support for substructural types as in
    \cite{DBLP:journals/corr/abs-2005-02247}.
  \item
    Extending the class of typing relations covered by
    the \ACode{\ACon{TypingTraversal}} abstraction and identifying
    a subclass for which it can be derived.
  \item
    Incorporating a decision procedure for equational
    substitution-lemmas based on $\sigma$-calculus as in Autosubst
    \cite{DBLP:conf/itp/SchaferTS15, DBLP:conf/cpp/StarkSK19}.
  \item
    Developing abstractions and derived lemmas beyond substition and
    type preservation, e.g. for confluence or denotational semantics
    as in \cite{DBLP:journals/pacmpl/AllaisA0MM18}.
  \end{itemize}

  \section{Conclusion}
  \label{sec:conclusion}

>>>>

@@

  The rest of this paper is structured as followed:
  Section~\ref{sec:syntax} introduces the multi-sorted syntax and compares it to the
  more common unsorted syntax.
  Section~\ref{sec:maps} introduces multi-sorted substitutions and renamings, and
  an abstraction to unify them.
  Section~\ref{sec:composition} introduces composition of multi-sorted
  substitions and renamings, and an abstraction to unify all four
  compositions.
  Section~\ref{sec:types} shows how to introduce the notion of types
  for multi-sorted syntax and how to define typing contexts.
  Section~\ref{sec:typing} shows how to define a multi-sorted typing relation,
  and an abstraction to unify type preservation lemmas for renamings and substitutions.
  Section~\ref{sec:subjectreduction} introduces a small-step
  operational semantics for System F and proves subject reduction.
  Section~\ref{sec:reflection} describes the class of object languages covered by
  our reflection algorithm.
  Section~\ref{sec:casestudies} discusses our case studies.
  Section~\ref{sec:related} discusses related work.
  Section~\ref{sec:future} concludes with future work.

@@

@@

  % Frameworks for programming language metatheory can be broadly
  % categorized by representation of variable bindings and
  % representation of typing and scoping.

  % Variable bindings have been represented as de Bruijn indices, de
  % Bruijn levels, names, locally nameless, locally named, shifted names,
  % nominal techniques, (parameterized) higher order abstract syntax.

  % Typing and scoping can be both extrinsic, both intrinsic, or
  % extrinsically typed and intrinsically scoped.

  As the amount of related work is rather large, we focus on work that
  is closely related to ours and refer to other papers for the
  broader
  picture\cite{DBLP:conf/cpp/StarkSK19,DBLP:journals/pacmpl/AllaisA0MM18}. In
  particular, we discuss abstractions for renamings
  and substitutions, comparable extrinsically typed approaches, but also
  intrinsically typed approaches that could be used as a basis for
  multi-sorted syntax, which as mentioned before is intrinsically typed
  in disguise.

@@

  The kit abstraction for unifying renamings and substitutions
  appeared first in an unpublished manuscript by
  McBride\cite{unpublished:mcbride2005kits}, and later in
  Benton et al.\cite{DBLP:journals/jar/BentonHKM12}. 
  Wood and Atkey\cite{DBLP:journals/corr/abs-2005-02247} propose an
  extension to kits that supports substructural types via resource
  vectors.
  In all three cases kits are formulated for intrinsic typing
  and scenarios with polymorphism are not yet considered.

@@

  \emph{Autosubst 2}\cite{DBLP:conf/cpp/StarkSK19} is a standalone
  code generator, which translates second-order HOAS specifications into
  mutual inductive term sorts. Compared to Autosubst 1, it features
  mutually recursive object languages, intrinsically scoping, and
  vectorized substitutions.
  Compared to our work, the syntax they generate takes the form of
  what we described as \emph{unsorted syntax} in
  Section~\ref{sec:syntax}, i.e. different syntactic categories are
  described by different types with different amount of indices for
  variable counts. To eliminate the need for interaction lemmas, they
  define the notion of \emph{vectorized substitution}, which combines
  the individual substitutions by putting them in a vector.

  Both works do not provide machinery for type preservation and do not
  unify renaming and substitution (although the latter is listed as
  future work). We believe their amazing work of creating a decision
  procedure for vectorized substitutions should also translate to our
  setting with multi-sorted substitutions.

  % Abstract binding trees

@@

  Allais et al.\cite{DBLP:conf/cpp/Allais0MM17} propose a powerful
  abstraction for denotational semantics and semantic fusion lemmas.
  In later work\cite{DBLP:journals/pacmpl/AllaisA0MM18}, they use
  generic programming to instantiate this abstraction for a class of
  object languages comparable to ours.
  They demonstrate how both renamings and substitutions can be
  described as semantics, how the four composition lemmas follow from
  their generic fusion lemma, and also provide an abstraction to unify
  renamings and substitutions.
  They show how to use their framework for both intrinsic and
  extrinsic typing, but have not yet explored the realm of polymorphism.

  This work is orthogonal, but compatible to our approach.
  We can instantiate their system using the hidden intrinsic type
  system of our multi-sorted syntax, but unfortunately then their
  semantics abstraction would only work with terms, which are untyped
  with respect to the actual extrinsic type system.
  We think it is worthwile to explore if their semantics abstraction
  can also be generalized to multi-sorted typing relations.

  \section{Future Work}
  \label{sec:future}
  We want to focus our future work in the following directions:
  \begin{itemize}
  \item
    Enabling more object languages by allowing subterms with less
    restrictions on the form of \ACode{S}, allowing recursive
    occurences in container types like lists and finite functions, and
    adding support for substructural types as in
    \cite{DBLP:journals/corr/abs-2005-02247}.
  \item
    Extending the class of typing relations covered by
    the \ACode{\ACon{TypingTraversal}} abstraction and identifying
    a subclass for which it can be derived.
  \item
    Incorporating a decision procedure for equational
    substitution-lemmas based on $\sigma$-calculus as in Autosubst
    \cite{DBLP:conf/itp/SchaferTS15, DBLP:conf/cpp/StarkSK19}.
  \item
    Developing abstractions and derived lemmas beyond substition and
    type preservation, e.g. for confluence or denotational semantics
    as in \cite{DBLP:journals/pacmpl/AllaisA0MM18}.
  \end{itemize}
  \vspace{10mm}
